---
title: "Module 2 Assessment"
output: pdf_document
---

### Problem 2

Suppose a particular crime is committed in Jerry’s apartment. We’d like to know whether Newman is guilty of the crime. We are torn as to whether we think he is guilty: we think it’s equally likely that he is guilty or not guilty. Suppose that, in similar situations, we know that if a suspect is guilty, 85% of the time their finger prints are found at the scene, and, we know that if a suspect is not guilty, 30% of the time their finger prints are found at the scene.

**a)**  What is the probability that Newman’s finger prints are found at the scene? Round your answer to three decimal places. Enter your solution into variable p2.1.


```{r}
p_a <- .5
p_a_comp <- 1-p_a
p_b_given_a <- .85
p_b_given_a_comp <- .3
p_b <- p_b_given_a*p_a + p_b_given_a_comp*p_a_comp
p2.1 = round(p_b, 3)
```

**b)**  If Newman’s finger prints are found at the scene, how likely is it that he is guilty? Round your answer to three decimal places. Enter your solution into variable p2.2.

```{r}
p_a_given_b <- (p_b_given_a*p_a)/p_b
p2.2 = round(p_a_given_b, 3)
```

**c)** If Newman’s finger prints are not found at the scene, how likely is it that he is guilty? Round your answer to three decimal places. Enter your answer into variable p2.3.

```{r}
p2.3 = 0
```

### Problem 3

The game of Yahtzee is played with five fair dice. The goal is to roll certain
‘hands’, such as Yahtzee (all five dice showing the same number) or a full house
(three of a kind and two of a kind). In the first round of a player’s turn, the player
rolls all five dice. Based on the outcome of that roll, the player has a second and
third round, where he/she can then choose to re-roll any subset of the dice to get
a desired hand.

**a)** What is the probability of rolling a Yahtzee (all 5 dice showing the same number) on the first round? Round your answer to 4 decimal places. Enter your answer into variable `p3.1`.


```{r}
p3.1 = 6/(6^5)
```

**b)** Suppose that, on the second round, the dice are {2, 3, 4, 6, 6}. You decide to
re-roll both sixes in the third round. What is the probability that you roll
either a small straight (exactly four dice are in a row) or a large straight (exactly five
dice are in a row)? Round your answer to three decimal places. Enter your solution into variable `p3.2`. (Note that to be "in a row" means that the dice come up with numbers that could be arranged in a row. Examples of exactly 5 numbers in a row are "1,2,3,4,5" or  "2,3,4,5,6". You may want to look up small and large straights in Yhatzee if you are unfamilar with the game or confused by this terminology).


```{r}
p3.2 = round(20/36,3)
```

**c)** What is the probability of rolling a
small straight (exactly four dice in a row) on the first round? Round your answer to have three decimal places. Enter your solution into variable `p3.3`.


```{r}
# prob 4 in a row - prob 5 in a row - prob 6 in a row?
p3.3 = 0
```